Housing Market Institutions Drive Race and Ethnicity Differences in Energy Consumption

### 2015-04-04 ConferencePaper/DraftEdwardsWoods.Rnw
Commented out the data description that sounded like it was from the RASS data set and replaced it with RECS.  All the figures in the structure type data now have references and have been tuned.  Also, there is text describing the data in tables and figures.

*******Observed differences evident between different households can at least be partially explained by race and ethnicity composition.  That is, there

Introduction/Literature Review:

@@

\begin{itemize}
  \item Description on RECS data size number of observations sampling method etc.
  \item Scope of data  
  \item Weighting to account for stratified sampling
  \item We need more tables and graphics here.
\end{itemize}

Variable and Data Description: 
Residential Energy Consumption Survey national data 2009 (RECS) was utilitized for this portion of the project. 12083 observations are included in the data, of which 11395 are included in the structure-tenure model. Weights provided in the data set were utilized to account from sample stratification.
  Combining dwelling type (dwltype) and the binary response variable ownrent, Dwltype includes six categories as follows: single family detached house, townhouse (including duplex, or row house), small apartment complex, large apartment complex, and mobile home. Mobile homes are notable for having considerably fewer observations than other categories and were ultimately dropped from the model. 
  The model is based on the following variables: income, resident ages, education, rental assistence, and ethnicity and a variable for HUD complaints by state.  Householder race was interacted with income and  HUD complaints.  Logged, midpoints, are used for income. Education is broken into six categories describing highest education level completed by the head of the household: elementary, some high school, high school. Some college, Associate’s, Bachelor’s, Master;s, Professional, and Doctorate degree.  The ethnicity variable is from one to six: Native American, Asian/Pacific Islander, Black/African American, Hispanic/Latino, White/Caucasian, other, and more than one race.  
  Over twice as many households choose to own their homes compared to renters.  As well we see that the majority of households, more than all other categories combined, live in single family detached houses (See Appendix).  Behind single family houses, large apartment complexes were the most likely dwelling types chosen.  Notably, apartments in general, were more commonly rented than owned as compared to the opposite for each other dwelling type.
	With respect to respondent education attainment, percentages of each category (see appendix) do differ from what can be expected.  It is entirely likely that this is an example of the survey effect on selection of respondents, i.e. that individuals with higher education attainment are more likely to respond to mailed surveys.

@@

	Approximately 2.7 times as many households choose to own their homes compared to renters.  As well we see that the majority of households, more than all other categories combined, live in singe family detached houses (See Appendix).  Behind single family houses, large apartment complexes were the most likely dwelling types chosen.  Notably, apartments in general, were more commonly rented than owned as compared to the opposite for each other dwelling type.

	With respect to respondent education attainment, it can be noted that the percentages of each category (see appendix) do differ from that of what can be expected.  It is entirely likely that this is an example of the survey effect on selection of respondents, i.e. that individuals with higher education attainment are more likely to respond to mailed surveys.  Comparing to a cursory internet search for example, more respondents hold graduate degrees and bachelor’s degrees, and less high school diplomas and below than that percentage constructed for the state of California (PPIC).  It is also of minor interest to note the seventy-nine respondents who claim to live in mobile homes and hold graduate degrees.  While that combination of characteristics is plausible enough without further examination, it does bring into question the honesty of respondent’s self-reported information.  But, by constructing a variable equal to the number of housemates over the age of 65 and tabulating it against just mobile homes within the dwelling type variable, we note that the majority of respondents state that 1346 seniors live in mobile homes compared to 643 mobile homes without senior citizens, which is a considerably different trend than homes largely without seniors in the other dwelling type categories particularly detached houses, bearing in mind that this is relatively and the majority of seniors recorded in the data do not live in mobile homes. This may suggests that seniors are utilizing mobile homes as retirement homes, possibly within retirement communities and that would help to explain the minor perceived discrepancy in the education variable.  It is also commonly held that seniors may be more likely to respond to surveys, much like voting, and this would suggest why there are more respondents in mobile homes than in every other category of dwelling aside from home ownership and small complex ownership.  It can be noted from the RASS survey that, in fact, mobile homes had the highest response rate (RASS).

Lastly, one of our variables, actually two, are count variables that account for the number of children under the age of five and between six and eighteen respectively.  This variable was constructed with the intention of finding whether larger structures, in general, such as houses and townhouses, are chosen more often by individuals requiring more room or who are exhibiting more permanent nesting behavior, as in choosing to purchase a home with the intention of residing there for a long period of time and becoming well established.

@@

Table 
% \ref{tab:RaceVStruct}
shows the count of observations by Race/Ethnicty and structure after the basic data handling described above.  Please note the dominance of single family detached homes, shown as SFDetached in the table and non-Hispanic Caucasians, shown as FALSEWt in the table.  These counts are unweighted and do not take into account the stratefied sampling scheme used to collect the data.

@@

Table
% \ref{tab:RegionVStruct}
gives a further description of the data is distributed around the country.  Please note that both the Western and North Eastern area have a larger fraction of large apartment buldings, LgApartment in the table, than the other two regions.

@@

Figure
% \ref{fig:kWhbyOwnRace}
shows a smooth approximation of the distribution of annual household electricity use in kWh per year.  The columns indicates the varias self-identified race and the columns indicate if the the household self identifies as Hispanic and if the home is owned or rented. The blank cells indicate that there are two few observations to generate a reliable smooth approximation of electricity consumption.

Please note that households that rent tend to have lower annual electricity consumption and a smaller variance in that consumption.  This is normal given the large number of rented apartments which are typically smaller and of more uniform size than single family dewellings.

\begin{figure}
\begin{center}
\caption{Annual kWh by Rent/Own and Race}
\label{fig:kWhbyOwnRace}
<<echo=false,fig=TRUE>>=
library(ggplot2)
qplot(KWH, data = RECS, xlim=c(0,30000), geom = "density" )+facet_grid(Hispanic+KOWNRENT~Householder_Race)+ylab("")+xlab("Annual kWh")
@
\end{center}
\end{figure}

% Cut for now. 
% \begin{figure}
% \begin{center}
% \caption{Annual kWh by Income}
% \label{fig:kWhbyIncome}
% <<echo=false,fig=TRUE>>=
% # library(ggplot2)
% # 
% # qplot(Income,KWH, data = RECS, xlim=c(0,30000) )+facet_grid(Hispanic+KOWNRENT~Householder_Race)+geom_smooth()
% # 
% # qplot(Income,KWH, data = RECS, xlim=c(0,30000),facets =EthRace~KOWNRENT )+geom_boxplot()
% @
% \end{center}
% \end{figure}

% 
% \begin{figure}
% \begin{center}\label{fig:HHbyRace}
% <<echo=false,fig=TRUE>>=
% # qplot(NHSLDMEM,data=RECS,title="Household Size by Race and Ethnicity")+xlab("Household Size")+facet_grid(Hispanic~Householder_Race)+xlim(0,7)+geom_density()
% @
% \end{center}
% \end{figure}

@@

\begin{itemize}
  \item differences in structure type x
  \item difference in tenure x
  \item differences in energy star controlling for own rent
  \item general differences in own vs rent.  
  \item explain with split incentives story
  \item difference in age of structure x
  \item differences in hot water fuel x
  \item difference in heating fuel x
    \item difference in climate/location x
\end{itemize}

% <<echo=false, results=tex>>=
% # xtable(table(RECS$EthRace,RECS$ESWWAC),caption="EnergyStar Wall AC by Race and Ethnicity", label="tab:EnergyStarWall")
% 
% @

% 
% \begin{figure}
% \begin{center}
% \caption{EnergyStar Wall AC by Race and Ethnicity}
% \label{fig:EnergyStarWall}
% <<echo=false, fig=TRUE>>=
% # library(ggplot2)
% # ggplot(RECS,aes(ESWWAC))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race) 
% @
% \end{center}
% \end{figure}

% 
% <<echo=false, results=tex>>=
% # library(xtable)
% # xtable(table(RECS$EthRace,RECS$UR),caption="Race by Urban/Rural", label="tab:UrbRural")
% @

@@

\begin{figure}
\begin{center}
\caption{Urban vs Rural by Race and Ethnicity}
\label{fig:UrbanRural}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(UR))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race)+ylab("Fraction") +xlab("Urban vs Rural")
@
\end{center}
\end{figure}

Note that non-Hispanic Caucasian and Native households are more likely to reside in a rural area while those that identifiy as Hispanic are in each category almost always more likely to live in urban areas than those that don't self identify as Hispanic.  The connection to energy consumption is through the correlation between rural households having larger residences that are more likely to be single family detached houses.

\begin{figure}
\begin{center}
\caption{Water Service by Race and Ethnicity}
\label{fig:WaterService}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(FUELH2O))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race) +coord_flip()+xlab("Water Service Fuel")+ylab("Fraction")
@
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\caption{Heating Fuel by Race and Ethnicity}
\label{fig:HeatFuel}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(FUELHEAT))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race) +coord_flip()+xlab("Heating Fuel")+ylab("Fraction")
@
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\caption{Climate Type by Race and Ethnicity}
\label{fig:HeatFuel}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(Climate_Region_Pub))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race) +coord_flip()+xlab("Climate Type")+ylab("Fraction")
@
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\caption{Age of Structure by Rent/Own and Race/Ethnicity}
\label{fig:AgebyOwnRace}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(YEARMADE))+geom_density()+facet_grid(Householder_Race~Hispanic+UR) +xlab("Year Built")+ylab("Fraction")
@
\end{center}
\end{figure}

>>>>

@@

The 2009 Residential Energy Consumption Survey (RECS) was the main source of data. 12083 observations are included in the data, of which 11905 are used for most of the models. Because of the complex sampling scheme, which involves both cluster sampling and over sampling of odd cells, weights are used in all analysis.  The reason that the majority of the observations were dropped was because the households were living on premiss for free, neighter owning or renting the property.  We also dropped a few observations because they were undersampled and the post-stratification weights  were so large to as not be credible.

The RECS data set contains information on both the dwelling structure type,  and an indicator if they own the property.  The dwelling types are divided into the six common categories: single family detached house, single family attached (including duplexs and row house, small apartment complexs, large apartment complexs, and mobile homes.

The RECS survey contains an enormous number of variables, 931, for each household including: income, resident ages, education, rental assistence, appliance and building characteristics, thermostat setpoints, householder ethnicity and race.  Income was recorded in categories but midpoints of the categories were used for most analysis. 
% 
% Over twice as many households choose to own their homes compared to renters.  As well we see that the majority of households, more than all other categories combined, live in single family detached houses (See Appendix).  Behind single family houses, large apartment complexes were the most likely dwelling types chosen.  Notably, apartments in general, were more commonly rented than owned as compared to the opposite for each other dwelling type.
% 
% With respect to respondent education attainment, percentages of each category (see appendix) do differ from what can be expected.  It is entirely likely that this is an example of the survey effect on selection of respondents, i.e. that individuals with higher education attainment are more likely to respond to mailed surveys.

@@

% 	Approximately 2.7 times as many households choose to own their homes compared to renters.  As well we see that the majority of households, more than all other categories combined, live in singe family detached houses (See Appendix).  Behind single family houses, large apartment complexes were the most likely dwelling types chosen.  Notably, apartments in general, were more commonly rented than owned as compared to the opposite for each other dwelling type.

% 	With respect to respondent education attainment, it can be noted that the percentages of each category (see appendix) do differ from that of what can be expected.  It is entirely likely that this is an example of the survey effect on selection of respondents, i.e. that individuals with higher education attainment are more likely to respond to mailed surveys.  Comparing to a cursory internet search for example, more respondents hold graduate degrees and bachelor’s degrees, and less high school diplomas and below than that percentage constructed for the state of California (PPIC).  It is also of minor interest to note the seventy-nine respondents who claim to live in mobile homes and hold graduate degrees.  While that combination of characteristics is plausible enough without further examination, it does bring into question the honesty of respondent’s self-reported information.  But, by constructing a variable equal to the number of housemates over the age of 65 and tabulating it against just mobile homes within the dwelling type variable, we note that the majority of respondents state that 1346 seniors live in mobile homes compared to 643 mobile homes without senior citizens, which is a considerably different trend than homes largely without seniors in the other dwelling type categories particularly detached houses, bearing in mind that this is relatively and the majority of seniors recorded in the data do not live in mobile homes. This may suggests that seniors are utilizing mobile homes as retirement homes, possibly within retirement communities and that would help to explain the minor perceived discrepancy in the education variable.  It is also commonly held that seniors may be more likely to respond to surveys, much like voting, and this would suggest why there are more respondents in mobile homes than in every other category of dwelling aside from home ownership and small complex ownership.  It can be noted from the RASS survey that, in fact, mobile homes had the highest response rate (RASS).
% 	
% Lastly, one of our variables, actually two, are count variables that account for the number of children under the age of five and between six and eighteen respectively.  This variable was constructed with the intention of finding whether larger structures, in general, such as houses and townhouses, are chosen more often by individuals requiring more room or who are exhibiting more permanent nesting behavior, as in choosing to purchase a home with the intention of residing there for a long period of time and becoming well established.

@@

Table 
 \ref{tab:RaceVStruct}
shows the count of observations by Race/Ethnicty and structure after the basic data handling described above.  Please note the dominance of single family detached homes, shown as SFDetached in the table and non-Hispanic Caucasians, shown as FALSEWt in the table.  These counts are unweighted and do not take into account the stratefied sampling scheme used to collect the data.

@@

Many of the structure type/ race and ethnicity combinations are relatively rare, e.g., Hispanic Asians living in mobile homes.  We have not collapsed any of these  categories.  Future research will look at collapsing some of the rare cells so that there are a more managable number of both structure types and race/ethnicity combinations.

Table
\ref{tab:RegionVStruct}
gives a further description of the data is distributed around the country.  Please note that both the Western and North Eastern area have a larger fraction of large apartment buldings, LgApartment in the table, than the other two regions.

@@

The table shows the range of climate zones and houseing law covered by the survey.  It also shows the reasonablness of our hypotheses that location and structure type choices are endogenous.

Figure
\ref{fig:kWhbyOwnRace}
shows a smooth approximation of the distribution of annual household electricity use in kWh per year.  The columns indicates the various self-identified race and the columns indicate if the the household self identifies as Hispanic and if the home is owned or rented. The blank cells indicate that there are too few observations to generate a reliable smooth approximation of electricity consumption.

Please note that households that rent tend to have lower annual electricity consumption and a smaller variance in that consumption.  This is normal given that rented apartments which are typically smaller and of more uniform size than single family dewellings.

\begin{figure}[htbp]
\begin{center}
\caption{Annual kWh by Rent/Own and Race}
\label{fig:kWhbyOwnRace}
<<echo=false,fig=TRUE>>=
library(ggplot2)
qplot(KWH, data = RECS, xlim=c(0,30000), geom = "density" )+facet_grid(Hispanic+KOWNRENT~Householder_Race)+ylab("")+xlab("Annual kWh")
@
\end{center}
\end{figure}

@@

@@

\begin{figure}[htbp]
\begin{center}
\caption{Urban vs Rural by Race and Ethnicity}
\label{fig:UrbanRural}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(UR))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race)+ylab("Fraction") +xlab("Urban vs Rural")
@
\end{center}
\end{figure}

Non-Hispanic Caucasian and Native American households are more likely to reside in a rural area while those that identifiy as Hispanic are, in each category, almost always more likely to live in urban areas than those that don't self-identify as Hispanic.  The connection to energy consumption is through the correlation between rural households having larger residences that are more likely to be single family detached houses.

Some of the large differences in those self-identifed hispanic categories are because of bery sparce cells.  There are only four Hispanic Pacific Islanders in the survey.

Both the water and space heating fuel show significant differences across race and ethnicity combinations.  The fuel choice something controled by the owner of the structure.  For households that rent this choice more likely to be electric heating and water service.  While electric heat and water service is often very expenseive on a month-to-month basis, the installation costs are typically quite low.  The installation cost is what is faced by the landlord.  The month-to-month cost is faced by the lessee.

Figure \ref{fig:HeatFuel} shows the distribution of heating fuel by race and ethnicity.  It is very clear that Natural Gas is the most common fuel for hot water service but there are some differences based on race and ethnicity.  Excluding the less common self-identified Hispanic categories, native americans and pacific islanders are more likely to have electric hot water service than non-Hispanic caucasions and households with a head of household that identifies as multi-racial is less likely to have electric hot water service.

\begin{figure}[htbp]
\begin{center}
\caption{Water Service by Race and Ethnicity}
\label{fig:WaterService}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(FUELH2O))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race) +coord_flip()+xlab("Water Service Fuel")+ylab("Fraction")
@
\end{center}
\end{figure}

Again, electric hot water service is more common in extream rural areas and rental housing.  Figure \ref{fig:HeatFuel} shows a similar figure for heating fuel.  Non-Hispanic African Americans and Native Americans are more likely to have electric heat than non-Hispanic caucasians, while Asian and Pacific Islanders have a large, ``Not applicable'', response category.   Again, Hispanic Caucasians are more likely to have electricity as their heating fuel than non-Hispanic Caucasians.

\begin{figure}[htbp]
\begin{center}
\caption{Heating Fuel by Race and Ethnicity}
\label{fig:HeatFuel}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(FUELHEAT))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race) +coord_flip()+xlab("Heating Fuel")+ylab("Fraction")
@
\end{center}
\end{figure}

Many of these different fuel choices can be partially explained by the locational choice made by the households.  Figure \ref{fig:HeatFuel} shows the climate type by race and ethnicity.  Non-Hispanic pacific islanders are more likely than non-hispanic caucaians to be in hot humid and marine environments and may explain the large number of not applicable responses for heating fuel.  Hispanic caucasions are also more likely to live in hot dry/mixed dry or hot humid climes that the other groups.  While these climate types will still have some heating load, heating behavior and heating fuel is less likely to be a strong determinant of total electricity use.  Heating is just less an issue than cooling.

\begin{figure}[htbp]
\begin{center}
\caption{Climate Type by Race and Ethnicity}
\label{fig:HeatFuel}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(Climate_Region_Pub))+geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]))+facet_grid(Hispanic~Householder_Race) +coord_flip()+xlab("Climate Type")+ylab("Fraction")
@
\end{center}
\end{figure}

There are also differences in the age of structure depending on both the race and ethnicity reported and if the household lives in an urban or rural area.  Figure \ref{fig:AgebyOwnRace} shows the distribution of the year the housing unit was built.  Columns indicate urban and rural households with the two left, $FALSE$, columns breaking out the non-Hispanic households and the right two, $TRUE$, showing the Hispanic households.  Rows are indicating the self-idendified race.  There are several blank cells, where a smooth histogram is not possible because of the small number of observations.

\begin{figure}[htbp]
\begin{center}
\caption{Age of Structure by Rent/Own and Race/Ethnicity}
\label{fig:AgebyOwnRace}
<<echo=false, fig=TRUE>>=
library(ggplot2)
ggplot(RECS,aes(YEARMADE))+geom_density()+facet_grid(Householder_Race~Hispanic+UR) +xlab("Year Built")+ylab("Fraction")+coord_flip()
@
\end{center}
\end{figure}

What should be most apparant is that rural houseing tends to be newer than urban househousing.  Urban households have a modal construction date in the 1980s while the rural areas are more likely to have a model construction date in the 2000s.

These diagrams should lead one to the conclusion that important variables, such as location and fuel choice differ significantly across race and ethnicity and should not be treated, without great consideration, as exogenous.
